Empirical observations let us infer that the advantages of the \ours\ architecture are rooted in its multi-rate hierarchical nature, as both the multi-rate sampling and the hierarchical interpolation complement the long-horizon forecasting task in \MLP-based architectures. In this ablation experiment, we quantitatively explore the effects and complementarity of the techniques in an \RNN-based architecture.

This experiment follows the Table~\ref{table:ablation_nhits_contributions} ablation study, reporting the average performance across \ETTm$2$, \Electricity, \Exchange, \TrafficL, and \Weather\ datasets. We define the following set of alternative models: \DilRNN$_{1}$, our proposed model with both multi-rate sampling and hierarchical interpolation, \DilRNN$_{2}$ only hierarchical interpolation, \DilRNN$_{3}$ only multi-rate sampling, \DilRNN\ with no multi-rate sampling or interpolation (corresponds to the original \DilRNN~\citep{chang2017dilatedRNN}).

Tab.~\ref{table:ablation_rnn} shows that the hierarchical interpolation technique drives the main improvements (\DilRNN$_{2}$), while the combination of both proposed components (hierarchical interpolation and multi-rate sampling) sometimes results in the best performance (\DilRNN$_{1}$), the difference is marginal. Contrary to the clear complementary observed in Table~\ref{table:ablation_nhits_contributions}, the \DilRNN\ does not improve substantially from the multi-rate sampling techniques. We find an explanation in the behavior of the \RNN\ that summarizes past inputs and the current observation of the series, and not the complete whole past data like the \MLP-based architectures. 

A key takeaway of this experiment is that \ours'\ hierarchical interpolation technique exhibits significant benefits in other architectures. Despite these promising results, we decided not to pursue more complex architectures in our work as we found that the interpretability of the \ours\ predictions and signal decomposition capabilities was not worth losing.

\input{tables/6-long/ablation_rnn}